\documentclass{zupan}

\begin{document}

\begin{exercise}{1}
  Prove that the empty set is a subset of every set.
\end{exercise}

\begin{solution}
  Let $A$ be a set. For all $x$, the condition $x \in \varnothing \Rightarrow x
  \in A$ is vacuously satisfied, so $\varnothing \subset A$.
\end{solution}

\begin{exercise}{2}
  A complex number $z$ is said to be algebraic if there are integers $a_0 ,
  \dots, a_n$, not all zero, such that \[a_0z^n + a_1z^{n - 1} + \cdots + a_{n
  - 1}z + a_n = 0.\] Prove that the set of all algebraic numbers is countable.
  \textit{Hint:} For every positive integer $N$ there are only finitely many
  equations with \[n + \abs{a_0} + \abs{a_1} + \cdots + \abs{a_n} = N.\]
\end{exercise}

\begin{exercise}{4}
  Is the set of all irrational real numbers countable?
\end{exercise}

\begin{solution}
  No, this set cannot be countable. Suppose that the set of all irrational real
  numbers $I = \mathbb{R} \setminus \mathbb{Q}$ is countable. Consider the set
  $\mathbb{Q} \cup I$. Since this set was constructed from the union of two
  countable sets, it must be countable. However, this is a contradiction since
  the real numbers are uncountable. Thus, the irrational reals cannot be
  countable.
\end{solution}

\begin{exercise}{22}
  A metric space is called \textit{seperable} if it contains a countable dense
  subset. Show that \(\mathbb{R}^k\) is seperable. \textit{Hint:} Consider the
  set of points which have only rational coordinates.
\end{exercise}

\begin{solution}
  We aim to show that \(\mathbb{R}^k\) contains a countable dense subset for
  any positive integer \(k\).
\end{solution}

\begin{exercise}{29}
  Prove that every open set \(R^1\) is the union of an at most countable
  collection of disjoint segments. \textit{Hint:} Use Exercise 22.
\end{exercise}

\end{document}
